---
title: "Kang IFN-β PBMC: MOFA-FLEX from R to MOFA2"
author: "Maximilian Nuber"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Kang IFN-β PBMC: MOFA-FLEX from R to MOFA2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment  = "#>",
    eval     = TRUE # heavy analysis; set eval=TRUE to run end-to-end
)
```

# Overview

This vignette demonstrates an end-to-end MOFA-FLEX workflow entirely in R:

1. **Load** the Kang et al. (2018) PBMC IFN-β stimulation dataset from
   `muscData`.
2. **Preprocess** with standard Bioconductor tools (scran / scuttle).
3. **Bridge** the resulting `SingleCellExperiment` — a subclass of
   `SummarizedExperiment` — into a Python `AnnData` via
   `sce_to_reticulate_anndata()`, using matrix-type-aware dispatch that
   converts the assay matrix to a NumPy array or a SciPy sparse matrix and
   applies zero-copy where possible.
4. **Train** a MOFA-FLEX model with `fit_mofaflex()`, saving the model to
   an HDF5 file compatible with MOFA2.
5. **Load** the saved model in the MOFA2 R package for downstream analysis.

## Required packages

```{r packages, message=FALSE, warning=FALSE}
# Bioconductor
library(muscData) # BiocManager::install("muscData")
library(scuttle) # BiocManager::install("scuttle")
library(scran) # BiocManager::install("scran")
library(SingleCellExperiment)

# MofaflexR (this package)
library(MofaflexR)

# Downstream — MOFA2
library(MOFA2) # BiocManager::install("MOFA2")
```

---

# Load data {#load}

`muscData::Kang18_8vs8()` provides 29,065 PBMCs from eight donors, each
split into a control and an IFN-β–stimulated condition (8 × 2 = 16 samples).

```{r load-data}
sce_full <- Kang18_8vs8()
sce_full
```

```
# class: SingleCellExperiment
# dim: 35,635 genes × 29,065 cells
# colData names: ind, stim, cluster, cell, multiplets
```

---

# Preprocess {#preprocess}

## Quality-control filtering

Remove multiplets and low-quality cells before normalisation.

```{r qc}
## Remove multiplets and cells without an assigned cell type
sce <- sce_full[, sce_full$multiplets == "singlet" & !is.na(sce_full$cell)]

## Keep cells with ≥ 200 detected genes
qc <- scuttle::perCellQCMetrics(sce)
sce <- sce[, qc$detected >= 200]

cat(sprintf("After QC: %d genes × %d cells\n", nrow(sce), ncol(sce)))
```

## Normalisation

Log-normalise using pooling-based size factors from scran.

```{r normalise}
set.seed(42)
# clusters <- scran::quickCluster(sce)
# sce      <- scran::computeSumFactors(sce, clusters = clusters)
sce <- scuttle::logNormCounts(sce)

assayNames(sce) # should include "logcounts"
```

## Select highly variable genes

Retain the 3,000 most variable genes to keep the model tractable.

```{r hvgs}
dec <- scran::modelGeneVar(sce)
hvgs <- scran::getTopHVGs(dec, n = 3000)
sce_hvg <- sce[hvgs, ]

cat(sprintf("Using %d HVGs from %d cells.\n", nrow(sce_hvg), ncol(sce_hvg)))
```

---

# Convert to AnnData {#anndata}

`sce_to_reticulate_anndata()` bridges the selected assay from the
`SingleCellExperiment` (a `SummarizedExperiment` subclass) to a Python
`anndata.AnnData` object using matrix-type-aware conversion:

* **Sparse matrices** (e.g. `dgCMatrix`) are converted to a matching SciPy
  sparse format (`csc_matrix` for CSC, `csr_matrix` for CSR, etc.) and
  zero-copy is used for the backing slot arrays where possible.
* **Dense matrices** (e.g. `dgeMatrix`, base `matrix`) are converted to NumPy
  arrays, with zero-copy attempted via the R-to-NumPy buffer protocol.
* All `colData` columns are forwarded to `adata.obs`; `rowData` goes to
  `adata.var`.
* The assay (originally features × cells) is transposed once to AnnData's
  expected `obs × vars = cells × features` layout.

The `logcounts` assay produced by `scuttle::logNormCounts()` is typically
stored as a sparse matrix, so the returned `adata$X` will be a SciPy sparse
matrix (CSR after transposition).

```{r convert}
adata <- sce_to_reticulate_anndata(
    sce_hvg,
    assay = "logcounts"
)
adata
```

The `stim` column in `adata$obs` (`"ctrl"` / `"stim"`) will be used by
MOFA-FLEX to split cells into two groups.

---

# Train a MOFA-FLEX model {#train}

Key choices:

* `group_by = "stim"` — split cells into **ctrl** and **stim** groups using
  the `stim` column in `obs`.
* `subset_var = NULL` — HVGs were already selected; skip the
  `highly_variable` filter.
* `save_path` — write the trained model to an HDF5 file.
* `mofa_compat = "full"` — save in MOFA2-readable format.
* View name will be **`rna`** (automatic when wrapping a single AnnData).

```{r train}
## Output path — adjust to a stable directory in your project
output_dir <- file.path("results", "mofaflex")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
model_path <- file.path(output_dir, "kang_mofaflex.hdf5")
model <- fit_mofaflex(
    data = adata,
    data_options = list(
        # group_by           = "stim",
        scale_per_group    = TRUE,
        subset_var         = NULL,
        plot_data_overview = FALSE
    ),
    model_options = list(
        n_factors    = 5L,
        weight_prior = "Horseshoe"
    ),
    training_options = list(
        max_epochs             = 5000L,
        early_stopper_patience = 50L,
        seed                   = 42L,
        device                 = "cuda",  # or "cpu" if no GPU is available
        save_path              = model_path
    ),
    mofa_compat = "full"  # write a MOFA2-readable HDF5 file
)
```

> **Compute time.** On a modern CPU this typically converges in a few minutes
> thanks to early stopping.

---

# Load the model in MOFA2 {#mofa2}

Because the model was saved with `mofa_compat = "full"`, the HDF5 file
contains all MOFA2-compatible weight and factor matrices.  Use the
`load_mofaflex_model()` wrapper (which handles minor HDF5-format differences
between **mofaflex** and MOFA2 automatically) to load it.

```{r mofa2-load}
mofa <- load_mofaflex_model(model_path)

## Attach sample metadata so colour_by aesthetics work.
## MOFA2 requires a 'sample' column matching the model's cell IDs.
cells_per_group <- MOFA2::samples_names(mofa)
meta <- do.call(rbind, lapply(names(cells_per_group), function(grp) {
    cells <- cells_per_group[[grp]]
    sdf <- as.data.frame(SummarizedExperiment::colData(sce_hvg)[cells, , drop = FALSE])
    data.frame(
        sample = cells, group = grp, sdf,
        check.names = FALSE, stringsAsFactors = FALSE
    )
}))
MOFA2::samples_metadata(mofa) <- meta
mofa
```

## Variance explained

```{r mofa2-r2, fig.cap="Variance explained per factor and group."}
MOFA2::plot_variance_explained(mofa, max_r2 = 0.1)
```

## Factor scores

```{r mofa2-factors, fig.cap="Factor scores coloured by stimulation condition."}
MOFA2::plot_factor(
    mofa,
    factors = 1:5,
    color_by = "stim",
    add_violin = FALSE,
    dodge = TRUE,
    add_boxplot = TRUE
)
```

## Top feature weights

```{r mofa2-weights, fig.cap="Top 10 feature weights for Factor 1."}
MOFA2::plot_top_weights(
    mofa,
    view      = "rna",
    factor    = 5,
    nfeatures = 10
)
```

```{r mofa2-umap, fig.cap="UMAP of MOFA-FLEX factor scores coloured by stimulation condition."}
factors <- MOFA2::get_factors(mofa, factors = 1:5)
df <- cbind(
    factors$group_1,
    as.data.frame(SummarizedExperiment::colData(sce_hvg)[rownames(factors$group_1), ])
)
umaps <- uwot::umap(factors$group_1, n_neighbors = 15, min_dist = 0.1, metric = "cosine")
df <- cbind(df, UMAP1 = umaps[, 1], UMAP2 = umaps[, 2])
ggplot2::ggplot(df, ggplot2::aes(x = UMAP1, y = UMAP2, color = stim)) +
    ggplot2::geom_point(size = 0.5, alpha = 0.6) +
    ggplot2::theme_minimal() +
    ggplot2::labs(title = "UMAP of MOFA-FLEX factors coloured by stimulation condition")
```

Alternatively, use basilisk to run MOFAFlex in a separate Python process. The model is saved to disk anyway and loaded with MOFA2. If `setBasiliskShared(TRUE)`, the Python model is loaded into the current R session.

```r 

basilisk::setBasiliskFork(TRUE) # use forked processes for parallelism (Linux/Mac only)
basilisk::setBasiliskShared(FALSE) # share the same Python environment across processes (saves memory)

output_dir <- file.path("results", "mofaflex")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
model_path <- file.path(output_dir, "kang_mofaflex.hdf5")

with_mofaflex_env({
  # sce_to_reticulate_anndata uses matrix-type-aware dispatch:
  # the logcounts assay (typically sparse) is converted to SciPy sparse;
  # zero-copy is used for the backing data arrays where possible.
  adata <- MofaflexR::sce_to_reticulate_anndata(sce_hvg, assay = "logcounts")
  MofaflexR::fit_mofaflex(
    data = adata,
    data_options = list(scale_per_group = TRUE, subset_var = NULL, plot_data_overview = FALSE),
    model_options = list(n_factors = 5L, weight_prior = "Horseshoe"),
    training_options = list(max_epochs = 5000L, early_stopper_patience = 50L, seed = 42L,
                            device = "cuda", save_path = model_path),
    mofa_compat = "modelonly"
  )
  return(NULL)
}, sce_hvg = sce_hvg, model_path = model_path)
```

---

# Session info {#session}

```{r session-info, eval=TRUE}
sessionInfo()
```
