---
title: "Bridging Single-Cell Multi-Omics: SCE and MAE to Python MuData"
author: "Maximilian Nuber"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Bridging Single-Cell Multi-Omics: SCE and MAE to Python MuData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.align  = "center",
  message    = FALSE,
  warning    = FALSE
)
```

# Introduction

**MofaflexR** provides zero-copy bridges between Bioconductor single-cell
objects and the Python `mudata` / `anndata` ecosystem. This vignette
demonstrates:

1. **`sce_to_reticulate_mudata()`**: Convert a
   `SingleCellExperiment` – with alternative experiments (altExps) encoding
   additional modalities such as surface-protein (ADT/CITE-seq) or
   chromatin-accessibility (ATAC) – into a `ReticulateMuData`.

2. **`mae_to_reticulate_mudata()`**: Convert a `MultiAssayExperiment` with
   multiple experiments into a `ReticulateMuData`.

3. **`ReticulateMuData` active bindings and methods**: Access Python MuData
   fields directly from R without any data copies.

All matrix transfers are **zero-copy**: R `dgCMatrix` objects are shared with
Python through `scipy.sparse.csc_matrix` pointers. No Python ↔ R
serialisation takes place unless you explicitly call `reticulate::py_to_r()`.

## Prerequisites

```{r check-env, eval = FALSE}
# Ensure the mofaflex_test conda environment is available (or your own)
reticulate::use_condaenv("mofaflex_test", required = TRUE)

# Or configure via ~/.Renviron:
#   RETICULATE_PYTHON=/path/to/python
```

```{r load-packages}
suppressPackageStartupMessages({
  library(MofaflexR)
  library(SingleCellExperiment)
  library(SummarizedExperiment)
  library(Matrix)
  library(S4Vectors)
})
```

---

# Synthetic CITE-seq Dataset

We construct a minimal **CITE-seq** dataset that faithfully mimics real
single-cell data: 300 cells, 1 000 genes (RNA), and 40 surface proteins (ADT).

```{r synthetic-data}
set.seed(42)

n_cells    <- 300L
n_genes    <- 1000L
n_proteins <- 40L

cell_ids    <- paste0("cell_",    seq_len(n_cells))
gene_ids    <- paste0("gene_",    seq_len(n_genes))
protein_ids <- paste0("protein_", seq_len(n_proteins))

# ---- RNA counts (sparse, ~5% density) -----------------------------------
rna_mat <- Matrix::rsparsematrix(
  nrow    = n_genes,
  ncol    = n_cells,
  density = 0.05,
  repr    = "C"
)
rna_mat <- round(abs(rna_mat) * 100)   # non-negative integer counts
rownames(rna_mat) <- gene_ids
colnames(rna_mat) <- cell_ids

# ---- ADT / protein counts (~50% density, lower diversity) ---------------
adt_mat <- Matrix::rsparsematrix(
  nrow    = n_proteins,
  ncol    = n_cells,
  density = 0.50,
  repr    = "C"
)
adt_mat <- round(abs(adt_mat) * 500)
rownames(adt_mat) <- protein_ids
colnames(adt_mat) <- cell_ids

# ---- Main SCE (RNA) -----------------------------------------------------
cell_types <- sample(c("CD4_T", "CD8_T", "B_cell", "Monocyte", "NK"),
                     n_cells, replace = TRUE)
batches    <- sample(c("batch1", "batch2"), n_cells, replace = TRUE)

sce <- SingleCellExperiment(
  assays  = list(counts = rna_mat),
  colData = DataFrame(
    cell_type = cell_types,
    batch     = batches,
    n_counts  = Matrix::colSums(rna_mat)
  ),
  rowData = DataFrame(
    gene_length = sample(500:5000, n_genes, replace = TRUE),
    highly_variable = sample(c(TRUE, FALSE), n_genes,
                             replace = TRUE, prob = c(0.1, 0.9))
  )
)

# ---- ADT altExp --------------------------------------------------------
adt_sce <- SingleCellExperiment(
  assays  = list(counts = adt_mat),
  rowData = DataFrame(
    isotype = sample(c("IgG", "IgM", "IgA"), n_proteins, replace = TRUE)
  )
)

SingleCellExperiment::altExp(sce, "adt") <- adt_sce

sce
```

```{r show-sce}
cat("Main SCE dimensions  : ", dim(sce)[1], "genes x", dim(sce)[2], "cells\n")
cat("altExp names         : ", altExpNames(sce), "\n")
cat("ADT dimensions       : ", dim(altExp(sce, "adt")), "\n")
```

---

# SCE → ReticulateMuData

## Convert with `sce_to_reticulate_mudata()`

`sce_to_reticulate_mudata()` maps the main SCE to one modality and each
`altExp` to its own modality. The conversion is **zero-copy**: all sparse
matrices remain in their original R memory locations.

```{r sce-to-mudata}
mdata <- sce_to_reticulate_mudata(
  x             = sce,
  main_modality = "rna",
  assay         = "counts",
  altexp_assay  = "counts"
)
mdata
```

`mdata` is a thin R6 proxy around a Python `mudata.MuData` object. All data
live in Python; R just holds a pointer.

## Active bindings – access Python fields directly

### `$shape` – dimensions

```{r shape}
mdata$shape          # c(n_obs, n_vars_total)
```

### `$obs_names` – shared observation names

```{r obs-names}
head(mdata$obs_names, 10)
length(mdata$obs_names)
```

### `$var_names` – all variable names

```{r var-names}
head(mdata$var_names, 6)
tail(mdata$var_names, 6)   # protein names appear at the end
```

### `$obs` – global observation metadata

After construction, `mudata.MuData` automatically pulls observation-level
metadata from the individual modalities into the global `obs` table.

```{r obs}
str(mdata$obs)
head(mdata$obs)
```

### `$mod` – extracted modalities

The `$mod` active binding returns a named R list of
`anndataR::ReticulateAnnData` wrappers.

```{r mod}
names(mdata$mod)
rna_adata <- mdata$mod$rna
rna_adata
```

Or equivalently using `[[`:

```{r bracket}
adt_adata <- mdata[["adt"]]
adt_adata
```

Both wrappers remain zero-copy: they point to the same Python `AnnData` objects
that are held inside the `MuData`.

### `$obsm` – multi-dimensional observation annotations

The `obsm` slot is empty initially; we add a synthetic PCA embedding to
illustrate the round-trip.

```{r obsm-assign}
# Create a synthetic PCA result (cells × 10 PCs)
pca_result <- matrix(rnorm(n_cells * 10), nrow = n_cells, ncol = 10,
                     dimnames = list(cell_ids, paste0("PC", 1:10)))

# Store directly via the Python AnnData binding
rna_adata$obsm[["X_pca"]] <- pca_result
rna_adata$obsm_keys()
```

After adding embeddings to a modality, call `mdata$update_obs()` to
synchronise the global obs index:

```{r update-obs}
mdata$update_obs()
cat("Global n_obs after update:", mdata$n_obs(), "\n")
```

### obs / var key introspection

```{r key-introspection}
mdata$obs_keys()     # global obs column names
mdata$var_keys()     # global var column names (if any)
mdata$uns_keys()     # unstructured annotation (empty here)
```

---

## Python method forwarding

Any Python `MuData` method can be called directly from R via `$<method>()`. 
The return value is automatically wrapped when appropriate.

### `$copy()` – deep copy

```{r copy}
mdata_copy <- mdata$copy()
inherits(mdata_copy, "ReticulateMuData")   # TRUE
identical(mdata_copy$py_mudata(), mdata$py_mudata())   # FALSE – new object
```

### `$update()` – sync global obs / var from modalities

```{r update}
mdata$update()
mdata$n_obs()
mdata$n_vars()
```

### `$py_call()` – generic Python method with auto-wrapping

```{r py-call}
# Any Python method not directly surfaced as an R method can be called via
# py_call().  The return is auto-wrapped into ReticulateMuData / ReticulateAnnData
# when applicable.
result <- mdata$py_call("copy")
class(result)
```

---

# MultiAssayExperiment → ReticulateMuData

## Construct a tiny MAE

```{r mae-build}
library(MultiAssayExperiment)

# Reuse the same RNA and protein matrices from above
sce_rna <- SingleCellExperiment(
  assays  = list(counts = rna_mat),
  colData = DataFrame(cell_type = cell_types, batch = batches)
)

sce_adt <- SingleCellExperiment(
  assays  = list(counts = adt_mat)
)

# Sample map: each experiment has identical primary ↔ colname
make_sm <- function(primary, colname, assay) {
  DataFrame(assay = assay, primary = primary, colname = colname)
}

smap <- rbind(
  make_sm(cell_ids, cell_ids, "rna"),
  make_sm(cell_ids, cell_ids, "adt")
)

col_df <- DataFrame(
  donor     = sample(paste0("donor_", 1:3), n_cells, replace = TRUE),
  row.names = cell_ids
)

mae <- MultiAssayExperiment(
  experiments = ExperimentList(rna = sce_rna, adt = sce_adt),
  colData     = col_df,
  sampleMap   = smap
)

mae
```

## Convert with `mae_to_reticulate_mudata()`

```{r mae-to-mudata}
mdata_mae <- mae_to_reticulate_mudata(mae, assay = "counts")
mdata_mae
```

```{r mae-inspect}
mdata_mae$modality_names()
mdata_mae$n_obs()
mdata_mae$n_vars()

# Check that MAE colData was propagated to global obs
str(mdata_mae$obs)
```

```{r mae-modalities}
rna_from_mae <- mdata_mae[["rna"]]
adt_from_mae <- mdata_mae[["adt"]]

cat("RNA modality: ", rna_from_mae$n_obs(), "obs x", rna_from_mae$n_vars(), "vars\n")
cat("ADT modality: ", adt_from_mae$n_obs(), "obs x", adt_from_mae$n_vars(), "vars\n")
```

---

# Real Dataset Examples (Optional)

The demonstrations above use synthetic data so the vignette renders offline
and instantly.  For a real CITE-seq workflow, replace the synthetic setup with
any SCE that has an ADT altExp.  Below is a sketch using the `scRNAseq`
package:

```{r real-dataset, eval = FALSE}
# ---- Using a real scRNAseq dataset ----------------------------------------
# Many scRNAseq datasets include altExps for spike-ins or protein panels.
# Example: Zilionis et al. mouse lung data (has ercc altExp)
#
# library(scRNAseq)
# sce_lung <- ZilionisLungData("mouse")
# altExpNames(sce_lung)   # check for altExps
#
# # Convert to MuData (use first assay of each altExp automatically)
# mdata_lung <- sce_to_reticulate_mudata(
#   x             = sce_lung,
#   main_modality = "rna",
#   assay         = "counts"
# )
# mdata_lung
```

For a `MultiAssayExperiment` from a real experiment:

```{r mae-real, eval = FALSE}
# ---- Using a MultiAssayExperiment from curatedTCGAData --------------------
# library(curatedTCGAData)
# mae_brca <- curatedTCGAData("BRCA",
#                             assays = c("RNASeq2GeneNorm", "miRNASeqGene"),
#                             dry.run = FALSE)
#
# mdata_brca <- mae_to_reticulate_mudata(mae_brca)
# mdata_brca$modality_names()
```

---

# Interoperability Round-trip

The `ReticulateMuData` object can be passed directly to any Python function
that accepts a `MuData` via its `.py_mudata()` raw accessor:

```{r py-round-trip, eval = FALSE}
# ---- Pass directly to Python code -----------------------------------------
mu <- reticulate::import("mudata", convert = FALSE)

# Write to h5mu
mu$write_h5mu(mdata$py_mudata(), "/tmp/my_mdata.h5mu")

# Read back – returns a Python MuData
py_loaded  <- mu$read_h5mu("/tmp/my_mdata.h5mu")
mdata_back <- ReticulateMuData$new(py_loaded)

mdata_back$obs_names
mdata_back$shape
```

---

# Session Info

```{r sessioninfo}
sessionInfo()
```
