---
title: "Matrix-aware R-to-Python bridge: SummarizedExperiment → NumPy / SciPy → AnnData"
author: "Maximilian Nuber"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zero-copy and near-zero-copy matrix bridges in MofaflexR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

This vignette explains how **MofaflexR** converts assay matrices from a
`SummarizedExperiment` (or any subclass, including `SingleCellExperiment`)
into Python NumPy / SciPy objects via `reticulate`, and then wraps the result
in an `anndata.AnnData` accessible from R via `anndataR::ReticulateAnnData`.

The conversion is **dispatch-based**: the package inspects the R class of the
assay matrix and selects the most appropriate Python representation, aiming to
avoid unnecessary copies.

Supported input matrix classes:

| R matrix class     | Python target                  | Zero-copy behavior |
|--------------------|--------------------------------|--------------------|
| `matrix`           | `numpy.ndarray`                | Zero-copy where possible via `r_to_py` buffer protocol |
| `dgeMatrix`        | `numpy.ndarray`                | Zero-copy via `@x` slot + Fortran-order reshape |
| `dgCMatrix`        | `scipy.sparse.csc_matrix`      | Zero-copy for `@x`, `@i`, `@p` slot arrays |
| `dgRMatrix`        | `scipy.sparse.csr_matrix`      | Zero-copy for `@x`, `@j`, `@p` slot arrays |
| `COO_SparseMatrix` | `scipy.sparse.coo_matrix`      | Zero-copy for values; coordinates require a copy |
| Other              | coerced to `csc_matrix` or `ndarray` | One R-level copy during coercion |

## Motivation

Single-cell count matrices can contain tens of millions of non-zero entries.
Copying them at the R ↔ Python boundary wastes memory and time.  Different
matrix classes expose memory in different ways; MofaflexR picks an appropriate
conversion for each:

* **Supported sparse formats** store their data, indices, and pointers in
  contiguous R integer/numeric vectors that can be handed to NumPy via the C
  buffer protocol (`reticulate::np_array()`), producing views with
  `flags.owndata == False`.
* **Dense base matrices and `dgeMatrix`** store their values as a single
  contiguous column-major vector which can be similarly exposed as a
  Fortran-order NumPy array.
* **COO coordinate matrices** require a 1→0 index-base conversion, so the
  coordinate arrays must be copied; only the value array can be a zero-copy
  view.

## How assay dispatch works

The container-level entry point is `sce_to_anndata()` (and its
`sce_to_reticulate_anndata()` wrapper).  Internally, assay conversion follows
this path:

```r
# Conceptual pseudocode
X <- se_assay_to_python_matrix(x, assay = assay, prefer_sparse = TRUE)
# X has shape (n_features, n_samples) — same as the R assay
X <- X$T   # transpose once: AnnData expects (obs=cells, vars=features)
```

The internal dispatcher selects converters by matrix class:

```r
# Simplified dispatch logic
if (is.matrix(x))                 return(.dense_matrix_to_numpy(x))
if (is(x, "dgeMatrix"))           return(.dense_matrix_to_numpy(x))
if (is(x, "dgCMatrix"))           return(.dgCMatrix_to_scipy_csc(x))
if (is(x, "dgRMatrix"))           return(.dgRMatrix_to_scipy_csr(x))
if (is(x, "COO_SparseArray"))     return(.coo_matrix_to_scipy_coo(x))
# fallback: coerce to dgCMatrix or densify, then convert
```

## Dense matrix conversion

### Base `matrix`

```r
# r_to_py on a base R matrix uses NumPy's buffer protocol.
# The result is a Fortran-order float64 ndarray; owndata = FALSE.
# (Observed behavior; the reticulate API does not formally guarantee it,
# but it is stable in practice across current versions.)
arr <- reticulate::r_to_py(mat)  # shape (nrow, ncol), owndata = FALSE
```

R matrices are column-major; NumPy's Fortran order matches this layout, so
no reordering is needed and no copy occurs during the conversion itself.

### `dgeMatrix`

`dgeMatrix` stores its values in a plain `@x` slot (a column-major numeric
vector), plus `@Dim`.  The conversion exposes `@x` as a 1-D zero-copy view,
then calls `.reshape((nr, nc), order = "F")`.  Because the 1-D buffer is
already laid out in column-major order, NumPy's reshape returns a
Fortran-contiguous 2-D **view** (`owndata = FALSE`); no copy is made.

## Sparse matrix conversion

### `dgCMatrix` → `scipy.sparse.csc_matrix`

A `dgCMatrix` with *m* features (rows) and *n* samples (columns) stores:

| Slot | Content | Length | R type |
|------|---------|--------|--------|
| `@x` | non-zero values | `nnz` | `double` → `float64` |
| `@i` | row indices (0-based) | `nnz` | `integer` → `int32` |
| `@p` | column pointers (`@p[j]` = start of col `j` in `@x`) | `n + 1` | `integer` → `int32` |

This maps directly to the CSC (Compressed Sparse Column) format.
`reticulate::np_array()` exposes each slot vector to NumPy without copying:

```r
# Pseudocode — zero-copy slot views
py_data    <- reticulate::np_array(mat@x, dtype = "float64")  # owndata = FALSE
py_indices <- reticulate::np_array(mat@i, dtype = "int32")    # owndata = FALSE
py_indptr  <- reticulate::np_array(mat@p, dtype = "int32")    # owndata = FALSE
csc <- sp$csc_matrix(
  reticulate::tuple(py_data, py_indices, py_indptr),
  shape = reticulate::tuple(nr, nc)
)
# SciPy stores references to the NumPy arrays; no further copy occurs.
```

### `dgRMatrix` → `scipy.sparse.csr_matrix`

`dgRMatrix` uses the CSR (Compressed Sparse Row) format.  The relevant slots are:

| Slot | Content | R type |
|------|---------|--------|
| `@x` | non-zero values | `double` |
| `@j` | column indices (0-based) | `integer` |
| `@p` | row pointers | `integer` |

These map directly to `csr_matrix((data, indices, indptr), shape=...)` with
the same zero-copy path as `dgCMatrix`.

### `COO_SparseMatrix` → `scipy.sparse.coo_matrix`

`COO_SparseMatrix` (from the `SparseArray` package) uses:

| Slot | Content | Notes |
|------|---------|-------|
| `@nzdata` | non-zero values | zero-copy `np_array` view |
| `@nzcoo`  | coordinate matrix, shape (nnz, 2), **1-based** | 1→0 conversion requires a copy |
| `@dim`    | `c(nrow, ncol)` | |

The coordinate arrays must be decremented from 1-based to 0-based before
being passed to SciPy.  This creates new R integer vectors (one copy each for
row and column), which are then turned into zero-copy NumPy views.  The value
array `@nzdata` itself is still a zero-copy view.

```r
# Coordinate conversion — unavoidable copy for 1→0 base change
row_0 <- coo@nzcoo[, 1L] - 1L   # new R vector (copy)
col_0 <- coo@nzcoo[, 2L] - 1L   # new R vector (copy)
py_row  <- reticulate::np_array(row_0, dtype = "int32")   # view of the new vector
py_col  <- reticulate::np_array(col_0, dtype = "int32")   # view of the new vector
py_data <- reticulate::np_array(coo@nzdata, dtype = "float64")  # zero-copy view
coo_py  <- sp$coo_matrix(
  reticulate::tuple(py_data, reticulate::tuple(py_row, py_col)),
  shape = reticulate::tuple(nr, nc)
)
```

## Orientation: why we transpose once

R assay matrices are stored features × samples (rows = features, columns =
samples/cells).  AnnData expects `obs × vars = cells × features`.  The
conversion therefore transposes the Python matrix exactly once:

```r
X <- se_assay_to_python_matrix(x, assay = assay)  # (features, cells)
X <- X$T                                            # (cells, features) — zero-copy for sparse
```

For SciPy sparse matrices, `.T` swaps the stored arrays' interpretation
(CSC ↔ CSR swap for example) without reordering values; `owndata` remains
`False`.  For NumPy dense arrays, `.T` returns a transposed view sharing the
same buffer.

## What CAN force a copy

| Trigger | Reason |
|---------|--------|
| Assay is an unsupported class (coercion fallback) | `as(mat, "dgCMatrix")` allocates a new R object |
| Assay is stored as a raw dense `integer` matrix | `storage.mode(x) <- "double"` makes a copy |
| COO coordinate arrays | 1 → 0 base conversion requires new integer vectors |
| `obs` / `var` metadata (pandas `DataFrame`) | R columnar `data.frame` → Python row-oriented; one copy per column, unavoidable |
| `py_to_r(adata$X)` | reticulate converts sparse → dense R matrix |
| `adata.X.toarray()` or `.todense()` in Python | explicit densification |
| Any `.copy()` call on a SciPy or NumPy object | explicit copy |

## Lifetime caveat

Python objects created via the zero-copy path hold **borrowed C references**
to the underlying R vectors.  This applies to any matrix class whose slot
vectors are exposed via `reticulate::np_array()` — not only `dgCMatrix`.

Ensure that:

1. The source R matrix object (and therefore its slots) remains alive and
   referenced by R while Python is using the converted object.
2. No R code triggers copy-on-modify on the same vector (standard R
   copy-on-modify semantics normally protect against this for unmodified
   objects).
3. The R garbage collector has not freed the underlying memory.

In typical usage (SCE / SE → AnnData → model fit → results back to R in the
same session), this is always satisfied as long as `sce` or equivalent is
held in the R global environment.

## Examples

```{r example-sparse, eval=FALSE}
library(MofaflexR)
library(SingleCellExperiment)
library(Matrix)

set.seed(1)

# ---- sparse dgCMatrix example -------------------------------------------
counts <- rsparsematrix(500, 100, density = 0.05, repr = "C")
counts <- methods::as(abs(round(counts)), "dgCMatrix")
sce    <- SingleCellExperiment(assays = list(counts = counts))
colnames(sce) <- paste0("cell", seq_len(ncol(sce)))
rownames(sce) <- paste0("gene", seq_len(nrow(sce)))

# Low-level: CSC matrix (features x cells orientation)
csc <- sce_assay_to_scipy_csc(sce)
cat("owndata:", reticulate::py_to_r(csc$data$flags$owndata), "\n")
# owndata: FALSE  (zero-copy view)

# Full AnnData via the generalized bridge (cells x genes)
rada     <- sce_to_reticulate_anndata(sce)
py_adata <- rada$py_anndata()
sp       <- reticulate::import("scipy.sparse", convert = FALSE)
cat("Shape:  ", unlist(reticulate::py_to_r(py_adata$shape)), "\n")  # 100 500
cat("Sparse: ", reticulate::py_to_r(sp$issparse(py_adata$X)), "\n") # TRUE
cat("owndata:", reticulate::py_to_r(py_adata$X$data$flags$owndata), "\n") # FALSE
```

```{r example-dense, eval=FALSE}
library(MofaflexR)
library(SummarizedExperiment)

# ---- dense dgeMatrix example (e.g. a PCA embedding stored as an assay) ---
set.seed(2)
dense_mat <- matrix(rnorm(500 * 50), nrow = 500L, ncol = 50L)  # 500 features x 50 cells
se <- SummarizedExperiment(assays = list(embedding = dense_mat))

# sce_to_anndata accepts any SummarizedExperiment subclass.
# The dense assay is converted to a NumPy ndarray via the buffer protocol.
adata <- sce_to_anndata(se, assay = "embedding")

np <- reticulate::import("numpy", convert = FALSE)
cat("Shape:     ", unlist(reticulate::py_to_r(adata$shape)), "\n")  # 50 500
cat("Is ndarray:", reticulate::py_to_r(np$ndim(adata$X)) == 2L, "\n") # TRUE
```
