---
title: "Zero-copy sparse bridge: SingleCellExperiment → scipy.sparse → AnnData"
author: "Maximilian Nuber"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zero-copy sparse bridge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

This vignette explains how **MofaflexR** converts a `SingleCellExperiment`
(SCE) assay stored as a `Matrix::dgCMatrix` into a Python
`scipy.sparse.csc_matrix` — without copying the underlying data arrays —
and then wraps it in an `anndata.AnnData` object accessible from R via
`anndataR::ReticulateAnnData`.

## Motivation

Single-cell RNA-seq count matrices can contain tens of millions of non-zero
entries.  Copying such matrices at the R ↔ Python boundary wastes both memory
and time.  MofaflexR avoids this by using `reticulate::np_array()`, which
exposes R's contiguous memory to NumPy via the **C buffer protocol** without
allocating new heap storage.

## Background: dgCMatrix memory layout

A `dgCMatrix` with *m* genes (rows) and *n* cells (columns) stores:

| Slot | Content | Length | R type |
|------|---------|--------|--------|
| `@x` | non-zero values (column-major) | `nnz` | `numeric` (double, 8 bytes) |
| `@i` | row indices of each `@x` entry (0-based) | `nnz` | `integer` (int32, 4 bytes) |
| `@p` | column pointer: `@p[j]` = index into `@x` where col `j` starts | `n + 1` | `integer` (int32, 4 bytes) |

This is exactly the CSC (Compressed Sparse Column) format used by
`scipy.sparse.csc_matrix`.

## How `sce_assay_to_scipy_csc()` works

```r
# Pseudocode (simplified)
sp         <- reticulate::import("scipy.sparse", convert = FALSE)
py_data    <- reticulate::np_array(mat@x, dtype = "float64")  # no copy
py_indices <- reticulate::np_array(mat@i, dtype = "int32")    # no copy
py_indptr  <- reticulate::np_array(mat@p, dtype = "int32")    # no copy
shape      <- reticulate::tuple(nr, nc)
csc        <- sp$csc_matrix(reticulate::tuple(py_data, py_indices, py_indptr),
                             shape = shape)
```

`reticulate::np_array()` uses the C buffer protocol to hand a pointer to the
existing R memory to NumPy.  The resulting arrays have
`flags.owndata == False`, confirming they are views, not copies.

## Orientation: why we transpose

`dgCMatrix` is genes × cells (R convention: rows are features).
AnnData uses obs × vars = cells × genes.
We call `csc.T` after construction:

```python
# csc.T is a csr_matrix of shape (cells, genes)
# scipy computes this by swapping the CSC ↔ CSR interpretation;
# no values are re-ordered — flags.owndata remains False.
```

## What CAN force a copy

| Trigger | Reason |
|---------|--------|
| Assay is not `dgCMatrix` on entry | `as(mat, "dgCMatrix")` may allocate |
| `py_to_r(adata$X)` | reticulate converts sparse → dense R matrix |
| `adata.X.toarray()` in Python | explicit densification |
| Pandas `DataFrame` for obs/var | unavoidable: R `data.frame` → Python copy |
| `scipy.sparse.csc_array` conversion | `.tocsc()` sorts & may copy indices |

## Lifetime caveat

The Python `csc_matrix` holds a **borrowed pointer** to R memory.  Ensure:

1. The source `dgCMatrix` remains alive (referenced by R) while Python uses it.
2. No R code triggers copy-on-modify for the same vector.
3. The R garbage collector has not freed the object.

In typical single-session usage (SCE → AnnData → model fit → result back to R)
this is always satisfied.

## Example

```{r example, eval=FALSE}
library(MofaflexR)
library(SingleCellExperiment)
library(Matrix)

set.seed(1)
counts <- rsparsematrix(500, 100, density = 0.05, repr = "C")
counts <- abs(round(counts))
sce    <- SingleCellExperiment(assays = list(counts = counts))
colnames(sce) <- paste0("cell", seq_len(ncol(sce)))
rownames(sce) <- paste0("gene", seq_len(nrow(sce)))

# --- zero-copy CSC matrix (genes x cells) ---
csc <- sce_assay_to_scipy_csc(sce)
cat("owndata:", reticulate::py_to_r(csc$data$flags$owndata), "\n")
# owndata: FALSE

# --- full AnnData (cells x genes) ---
rada     <- sce_to_reticulate_anndata(sce)
py_adata <- rada$py_anndata()
sp       <- reticulate::import("scipy.sparse", convert = FALSE)
cat("Shape:   ", unlist(reticulate::py_to_r(py_adata$shape)), "\n")
cat("Sparse:  ", reticulate::py_to_r(sp$issparse(py_adata$X)), "\n")
cat("owndata: ", reticulate::py_to_r(py_adata$X$data$flags$owndata), "\n")
# Shape:    100 500
# Sparse:   TRUE
# owndata:  FALSE
```
